---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r echo = FALSE}
rm(list=ls()) ### To clear namespace
library(knitr)
opts_chunk$set(fig.width=6, fig.height=4, fig.align = "center", digits = 2,
               echo=TRUE, results = TRUE, warning=FALSE, message=FALSE)
```
## Loading and preprocessing the data

We load the activity data, and convert the `date` column to class `POSIXct` so that we can later use the `weekdays` function on dates.

It appears that each interval was originally encoded by giving its start time as the number of hours since the time recorded as 0, followed by the number of minutes, with no space or punctuation.  Fo example the number 135 would represent ''1 hour, 35 minutes since start time.''  For easier plotting, we edit the intervals column so that the intervals are simply numbered in order, starting at 0.  So for example, interval 5 is the 4th interval, which takes place from 15-20 minutes from start time.

```{r}
data <- read.csv("activity.csv", header = TRUE, na.strings = "NA", 
                 colClasses = c("integer", "POSIXct", "integer"))
convert <- function(time){
    return ((time %/% 100 * 60 + time %% 100)/5)
}
data$interval <- convert(data$interval)
```

## What is mean total number of steps taken per day?

First, we will attempt to answer this question ignoring all missing values.  We use `tapply` to obtain a vector `by_day` which gives total number of steps on each day, and then find the mean
and median number of steps per day.

```{r}
by_day <- tapply(data$steps, data$date, sum, na.rm = TRUE)
mean(by_day)
median(by_day)
```

The average number of steps per day is approximately `r round(mean(by_day), 2)` and the median is `r median(by_day)`.

```{r}
library(ggplot2)
df <- data.frame(by_day)
ggplot(df, aes(df$by_day)) +
           geom_histogram(binwidth = 500, color = "red4", fill = "red3") +
           labs(x = "Steps", y = "Number of Days") +
           ggtitle("Histogram of Total Steps per Day")
```

## What is the average daily activity pattern?

```{r}
library(dplyr)
df <- data %>% group_by(interval) %>% summarize(steps = mean(steps, na.rm = TRUE))
ggplot(df, aes(interval, steps)) +
    geom_line(color = "red4") +
    labs(x = "Five-minute Interval", y = "Average Number of Steps") +
    ggtitle("Average Number of Steps Per Interval")
```

We calculate the interval with the highest average number of steps
```{r}
    max_interval <- df$interval[which.max(df$steps)]
```

To make this answer more meaningful, we can convert it to hours and minutes:
```{r}
    ## compute number of hours
    (max_interval * 5) %/% 60
    ## compute number of minutes
    (max_interval * 5) %% 60
```
The highest average number of steps occurs in the 5-minute interval that starts `r (max_interval * 5) %/% 60` hours and `r (max_interval * 5) %% 60` minutes after 00:00.

## Imputing Missing Values

```{r}
missing <- is.na(data$steps)
## compute the number of missing values
sum(missing)
```
There are `sum(missing)` missing values in our data set.  We replace each missing value by the average number of steps for the corresponding interval, which we can obtain from the dataframe `df` constructed above.
```{r}
for(idx in 1:dim(data)[1]){
    if(is.na(data$steps[idx])){
        my_interval <- which(df$interval == data$interval[idx])
        data$steps[idx] <- df$steps[my_interval]
        }
    }
```

We then repeat the computations to find the average and median number of steps per day.

```{r}
by_day <- tapply(data$steps, data$date, sum)

## compute the mean:
mean(by_day)
## compute the median:
median(by_day)
```

Imputing the missing values cuases both the mean and the median to increase; but the mean increases more.  Without the missing values, the mean was significantly smaller than the median.

```{r}
df <- data.frame(by_day)
ggplot(df, aes(df$by_day)) +
    geom_histogram(binwidth = 500, color = "red4", fill = "red3") +
    labs(x = "Steps", y = "Number of Days") +
    ggtitle("Total Steps per Day (Missing Data Imputed)")
```
From the graph, it appears that the large number of days with zero steps in our original dataset was due to missing data; perhaps corresponding to days when the subject did not carry their device.

## Are there differences in activity patterns between weekdays and weekends?

We create a factor variable which records whether each date is a weekday or weekend, and plot the number of steps in each interval averaged over all weekdays, and over all weekends.

```{r}
weekend <- function(day){
    weekdays(day) %in% c("Saturday", "Sunday")
}

data$weekend <- factor(weekend(data$date), labels = c("Weekday", "Weekend"))
library(dplyr)
my_data <- data %>% group_by(weekend, interval) %>% summarize(meansteps = mean(steps))
ggplot(data = my_data, aes(x = interval, y = meansteps)) + geom_line(color = "red4") +
    facet_grid(weekend ~ .) +
    labs(x = "Five-Minute Interval", y = "Average Number of Steps") + 
    ggtitle("Average Steps Per Five-Minute Interval")
```
Assuming the start time for daily measurements occurs at midnights, it appears that the subject is less active early in the morning on weekends, than on weekdays; but more active in the afternoon and evening.  Assuming a typical work or school schedule, this is a fairly believable finding.